---
title: "NYPD Shooting Data Analysis"
author: "George B. Wofford"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
params:
  data_url: "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(lubridate)
library(dplyr)
library(ggplot2)
```

# Introduction
This report explores publicly available NYPD shooting incident data from New York City. The primary goal is to understand patterns and trends in shootings over time and across boroughs, and to investigate factors that may influence the likelihood of a shooting resulting in a murder classification.

## Key Questions:

* How have shootings changed over the years captured in the dataset?
* Are there differences in the number of shootings among the boroughs of NYC?
* Is the timing of shootings (time of day) related to their frequency?
* Can basic variables (year, hour, borough) help us model the likelihood that a shooting is classified as a murder?

By addressing these questions, we aim to gain insights into temporal and spatial trends, and to experiment with a simple predictive model to see if basic factors correlate with shootings being deadly. Throughout, we will remain cautious about our interpretations and discuss possible sources of bias.

# Data Source and Description

The data is sourced directly from the NYC Open Data Portal using the following URL provided as a parameter: https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD

This dataset includes information on shooting incidents in NYC, containing details such as the date, time, borough, victim and perpetrator demographics, and indicators of whether the incident was classified as a murder.

Note on Reproducibility:
This R Markdown is fully reproducible. It imports data directly from the provided URL, ensuring that anyone knitting this document with the same code and environment can recreate the analysis and visualizations.

```{r data_source}
data_url <- params$data_url
nypd_shootings <- read_csv(data_url)
```

# Data Preparation and Cleaning
Before analysis, we need to ensure the data is in a tidy and appropriate format. We will:

* Convert date and time columns into proper Date and time objects.
* Convert categorical variables such as borough and demographic groups into factors.
* Remove columns not needed for this analysis (e.g., unique incident keys, coordinate columns) to simplify the dataset.

```{r tidy}
# Convert date from character to Date
nypd_shootings$OCCUR_DATE <- as.Date(nypd_shootings$OCCUR_DATE, format = "%m/%d/%Y")

# Convert time from character to a time object
nypd_shootings$OCCUR_TIME <- hms(nypd_shootings$OCCUR_TIME)

# Convert several categorical variables to factors
nypd_shootings$BORO <- as.factor(nypd_shootings$BORO)
nypd_shootings$LOC_OF_OCCUR_DESC <- as.factor(nypd_shootings$LOC_OF_OCCUR_DESC)
nypd_shootings$PERP_AGE_GROUP <- as.factor(nypd_shootings$PERP_AGE_GROUP)
nypd_shootings$PERP_SEX <- as.factor(nypd_shootings$PERP_SEX)
nypd_shootings$PERP_RACE <- as.factor(nypd_shootings$PERP_RACE)
nypd_shootings$VIC_AGE_GROUP <- as.factor(nypd_shootings$VIC_AGE_GROUP)
nypd_shootings$VIC_SEX <- as.factor(nypd_shootings$VIC_SEX)
nypd_shootings$VIC_RACE <- as.factor(nypd_shootings$VIC_RACE)
nypd_shootings$JURISDICTION_CODE <- as.factor(nypd_shootings$JURISDICTION_CODE)

# Remove unnecessary columns (coordinates and largely missing columns)
nypd_shootings <- subset(nypd_shootings, 
                         select = -c(INCIDENT_KEY, 
                                     LOC_OF_OCCUR_DESC, 
                                     X_COORD_CD, 
                                     Lon_Lat, 
                                     Latitude, 
                                     Longitude))

summary(nypd_shootings)
```

# Exploratory Data Analysis and Visualizations
## Shootings Over the Years
We first explore how shootings vary by year to detect any long-term trends.

```{r analyze_shootings_by_year}
nypd_shootings <- nypd_shootings %>% mutate(YEAR = year(OCCUR_DATE))

yearly_counts <- nypd_shootings %>%
  group_by(YEAR) %>%
  summarize(Total_Shootings = n())

ggplot(yearly_counts, aes(x = YEAR, y = Total_Shootings)) +
  geom_col(fill = "steelblue") +
  labs(title = "Number of Shootings per Year",
       x = "Year",
       y = "Number of Shootings") +
  theme_minimal()
```

**Interpretation**: Some years may have more shootings than others, suggesting underlying shifts in policy, socioeconomic conditions, or law enforcement practices.

## Shootings by Borough
Next, we look at the distribution of shootings across NYC boroughs to see which areas have higher frequencies.

```{r analyze_shootings_by_boro}
boro_counts <- nypd_shootings %>%
  group_by(BORO) %>%
  summarize(Total_Shootings = n())

ggplot(boro_counts, aes(x = BORO, y = Total_Shootings)) +
  geom_col(fill = "darkred") +
  labs(title = "Number of Shootings by Borough",
       x = "Borough",
       y = "Number of Shootings") +
  theme_minimal()
```

**Interpretation**: Certain boroughs show consistently higher shooting counts. This might raise questions about neighborhood-level differences or other structural factors.

## Shootings by Time of Day
We also consider the time of day. Do shootings peak at certain hours?

```{r analyze_shootings_by_time}
nypd_shootings <- nypd_shootings %>% mutate(HOUR = hour(OCCUR_TIME))

hour_counts <- nypd_shootings %>%
  group_by(HOUR) %>%
  summarize(Total_Shootings = n())

ggplot(hour_counts, aes(x = HOUR, y = Total_Shootings)) +
  geom_line(color = "purple") +
  geom_point(color = "purple") +
  labs(title = "Number of Shootings by Hour of Day",
       x = "Hour of Day (24-hour format)",
       y = "Number of Shootings") +
  theme_minimal()
```

**Interpretation**: If we see a pattern, perhaps increased shootings late at night, it might suggest that certain social activities or reduced police presence align with higher incident counts.

## Statistical Modeling
We will fit a simple logistic regression model to investigate if YEAR, HOUR, and BORO can help predict whether a shooting is classified as a murder (STATISTICAL_MURDER_FLAG).

```{r analyze_shootings_model}
# Ensure murder flag is a binary factor
nypd_shootings$STATISTICAL_MURDER_FLAG <- factor(nypd_shootings$STATISTICAL_MURDER_FLAG, 
                                                 levels = c(FALSE, TRUE))

# Fit a logistic regression model
murder_model <- glm(STATISTICAL_MURDER_FLAG ~ YEAR + HOUR + BORO, 
                    data = nypd_shootings, 
                    family = binomial(link = "logit"))

summary(murder_model)
```

**Interpretation**: The coefficients indicate which factors might be associated with higher odds of a shooting being classified as a murder. A positive coefficient for a given borough, for instance, might suggest that shootings in that borough have a higher probability of being deadly, controlling for year and hour. This is a simplistic model and should not be taken as definitive. Nonetheless, it demonstrates how data can be used to attempt predictive or explanatory modeling.

# Discussion and Potential Biases
## Reporting and Collection Bias:
The dataset only includes reported shootings, so it may not reflect all incidents. Reporting practices vary over time and between communities, influencing the apparent trends.

## Temporal Bias:
Over multiple years, changes in policies, economic conditions, and societal trends can affect both the frequency and classification of shootings. Without context, raw trends could be misinterpreted.

## Categorization and Missing Data:
Many variables were turned into factors, and some categories contain unknown or null values. Missing or imprecise data may skew patterns, especially if some groups are less frequently identified.

## Analyst Bias:
The selection of which variables to visualize and the simplicity of the chosen model reflect certain assumptions. A more thorough analysis might incorporate additional demographic details, spatial analysis using coordinates (if retained), or more robust modeling techniques.

# Conclusion
This exploratory analysis of NYPD shooting data provides a high-level look at how shootings vary over time, by borough, and by time of day. We introduced a simple logistic model to see if a few variables predict the classification of a shooting as a murder. The findings highlight temporal trends, geographic differences, and potential time-of-day patterns.

However, the analysis is only a starting point. It raises further questions about the underlying drivers of these patterns, and any interpretation must be cautious due to the potential biases in data collection and categorization. A deeper dive with more sophisticated modeling, external data sources, and domain expertise would be needed to draw stronger, more actionable conclusions.

Next Steps:

* Incorporate socioeconomic or law enforcement policy data for richer context.
* Consider more advanced models or machine learning approaches.
* Conduct sensitivity analyses to understand the impact of missing or uncertain data.

Overall, while we have visualized key trends and fit a basic model, we recognize that this analysis merely scratches the surface of understanding complex social phenomena like shootings in a large, diverse city.